Analysis of Major Crimes in Toronto And Strategies for Safety Amidst Increasing Numbers

### 2024-01-25 outputs/paper/paper.qmd
Change and move some content

---
title: "Analysis of Major Crimes in Toronto And Suggestions on How to Protect Ourselves"
author: "Boxuan Yi"
date: 22 January 2024
thanks: "Code and data in this analysis is available at: https://github.com/Elaineyi1/Analysis_of_Major_Crime_Indicators"
abstract: "Safety is one of the most important factors that people consider when deciding where to live and go. This report uses data from Open Data Toronto to analyze the occurrences of various types of crimes in different regions and time of Toronto, as well as the trend over the past ten years. The result shows that Downtown Toronto and evenings & noon have a relatively high occurrences of crimes, which could be used as a suggestion on both individual and government dicision-making."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

Major crimes statistics is an important tool for resource allocation and law enforcement strategy. It also provides the public a clear picture of the safety and helps individuals make informed decisions, influencing choices such as when to go out and where to buy a house. For this paper, I will use open-access data, titled ‘Major Crime Indicator’ from the OpenDataToronto [@rOpenDataToronto] to analyze the distribution of major crimes—classified as Assault, Break and Enter, Auto Theft, Robbery, and Theft Over—over the past ten years. My analysis will also delve into practical actions ordinary citizens and the government can do to enhance safety and efficiency.

@@

## Data Collection
The dataset contains all Major Crime Indicators occurrences by report data. The categories of major crimes are Assault, Break and Enter, Auto Theft, Robbery and Theft Over. The latest refresh was on Jan 11, 2024.

@@

## Data Analysis and Visualizations
The Major Crimes in Toronto dataset contains 27 columns and 372899 rows, including the repeated occurrences with different crime categories. All the samples of major crimes were reported from 2014 to 2023. Among 27 columns, the variables I will use are x_id, report_year, occ_year, occ_month, occ_dow (Day of the Week Offence Occurred), occ_hour, mci_category, division. I will create two more variables, division_area, which classifies 18 police divisions (including a NSA) into 6 regions of Toronto (including Other for NSA), and weekend_weekday, which indicates whether the occurrence happen on a weekday or a weekend.

```{r}
#| message: false
#| echo: false
#| label: tbl-clean_mci_data
#| tbl-cap: The Number of Major Crimes and the Month with the Most Crimes (2014-2023)
# Create a table with year, 2014 to 2023, the number of crimes in each year, and the month with the most crimes correspondingly.
clean_mci_data |>
  filter(occ_year >= 2014) |>
  group_by(occ_year, occ_month) |>
  summarise(n = n()) |>
  group_by(occ_year) |>
  summarise(
    total_number_of_crimes = sum(n),
    the_most_frequent_month = occ_month[which.max(n)]
  ) |>
  distinct(occ_year, .keep_all = TRUE) |>
  select(occ_year, total_number_of_crimes, the_most_frequent_month) |> 
  kable(
    col.names = c("Year", "Total Number of Crimes", "Month with the Most Crimes"),
    booktabs = TRUE
  )
```
Table 1 provides a brief summary of the total number of crimes from 2014 to 2023. Over this period, the overall number of crimes shows a consistent increase, except for the years 2020 and 2021, which have a decline probably due to the impact of the pandemic. However, in 2022 and 2023, there is a significant increase by more than 6000. In each year, the month with the highest number of crimes tends to be either July, August, or October, concentrating primarily in the summer-fall period.

@@

Figure 1 visualizes the annual distribution of major crime types with five categories: Assault, Break and Enter, Auto Theft, Robbery, and Theft Over. Across all years, 'Assault' records the highest count, followed by 'Break and Enter' and 'Auto Theft'. 'Robbery' and 'Theft Over' show comparatively lower numbers. Notably, there has been a slight decrease in “Break and Enter” since the pandemic, and an obvious increase in 'Auto Theft' during 2022 and 2023, leading ‘Auto Theft’ surpassing the number of 'Break and Enter’.

@@

```{r}
#| include: false
clean_mci_data <- clean_mci_data |> 
  mutate(
    weekday_weekend =
      case_match(
        occ_dow,
        "Wednesday" ~ "Weekday",
        "Tuesday" ~ "Weekday",
        "Thursday" ~ "Weekday",
        "Sunday" ~ "Weekend",
        "Monday" ~ "Weekday",
        "Friday" ~ "Weekday",
        "Saturday" ~ "Weekend",
        "" ~ "Other"))
clean_mci_data
```
There are 18 police divisions, which are D11, D12, D13, D14, D22, D23, D31, D32, D33, D41, D42, D43, D51, D52, D53, D54, D55, and NSA. I put them into 6 geographical regions, Etobicoke-York, Etobicoke, North York, Scarborough, Toronto-East York, and other, as shown in Table 2. Based on Figure 2, Toronto-East York reports the highest amount of crimes, and Etobicoke-York has the lowest. North York, Scarborough, and Etobicoke-York have similar numbers, but the rate of increase in North York is higher than that observed in the other two. 
```{r}
#| message: false
#| echo: false
#| label: tbl-division_area
#| tbl-cap: Police divisions in 6 areas(including other)

@@

```{r}
#| message: false
#| echo: false
#| label: fig-area_of_crimes
#| fig-cap: Total Number of Crimes in Different Police Division Areas
clean_mci_data |>
  filter(occ_year >= 2014) |>
  count(occ_year, division_area) |>
  ggplot(aes(x = occ_year, y = n, color = division_area)) +
  geom_point() +
  geom_jitter() +
  geom_smooth(method = lm, se = FALSE) +
  labs(x = "Year",
       y = "Total Number of Crimes",
       color = "Division Area")
```

```{r}
#| message: false
#| echo: false
#| label: fig-time_of_crimes
#| fig-cap: Number of Crimes in a Day
clean_mci_data |>
  filter(weekday_weekend == 'Weekday' | weekday_weekend == 'Weekend') |>
  count(occ_hour, weekday_weekend) |>
  ggplot(mapping = aes(x = occ_hour, y = n, color = weekday_weekend)) +
  geom_line() +
  theme(legend.position = "right") +
  labs(x = "Time In a Day", y = "Number of Crimes", color = "Weekday or Weekend") +
  scale_color_manual(values = c("Weekday" = "steelblue3", "Weekend" = "indianred3"))
```
\newpage
Figure 3 indicates the preferred time to commit crimes. The graph reveals a higher likelihood of criminal behaviours during the evening, with occurrences steadily increasing as the night progresses towards midnight and afterwards, the occurrences decrease until six in the morning. Interestingly, there is a surprising local peak at noon, while 11:00 and 13:00 do not show a high probability.

@@

# Conclusion
In summary, ‘Assault’ has always ranked as the most common major crime over the past ten years. However, the difference between ‘Auto Theft’ and ‘Assault’ is shrinking, as the number of ‘Auto Theft’ is increasing rapidly. There need be more safety actions on automobiles, and people should lock the doors and not leave any packages visible through car windows to improve automobile security. Due to the difficulty to prove ‘Assault’ as a crime, installing cameras at the front door or home security systems is a good idea to provide evidence for ‘Assault’, while it can also record suspicious individuals and vehicles near one's residence. Taking proactive steps, such as responding to knocks cautiously while keeping the door locked, and refraining from touching anything until the police have inspected, is crucial for personal safety and should be instilled in children as well.

@@

Focus on safety during midnight and noon hours is important too. The most common time for a crime occurrence include noon and midnight, especially between 6:00 PM to 8:00 PM. In order to prevent such incidents, people who work from home but go out for lunch, as well as those with children returning home for lunch around noon, should keep their ears and eyes open for suspicious surroundings. Evening hours in any region also need more cautions from residents and security from the government.

It is crucial to recognize that, though Toronto-East York records the most crimes, it also has the largest number of residents. Therefore, it could be biased to claim that Toronto-East York has the highest crime rate, or North York is getting dangerous, as North York has witnessed a growing population in recent years. Moreover, when creating this dataset, one occurrence number may have a few rows with different major crime indicators types used to categorize the occurrence. Therefore, some incidents of 'Auto Theft,' 'Robbery,' and 'Theft Over' may be potentially recorded multiple times under 'Break and Enter.' Consequently, the number of crimes classified as just ‘Break and Enter’ is overestimated, the actual number may be lower than what is shown now in the dataset.

>>>>

---
title: "Analysis of Major Crimes in Toronto And Strategies for Safety Amidst Increasing Numbers"
author: "Boxuan Yi"
date: 22 January 2024
thanks: "Code and data in this analysis is available at: https://github.com/Elaineyi1/Analysis_of_Major_Crime_Indicators"
abstract: "Safety is one of the most important factors that people consider when deciding where to live and go. This paper uses open-access data to analyze the occurrences of various types of crimes in Toronto and their trends over the past ten years. Crimes are classified based on regions where they happen and the times when they occur to further analyze the safety issue. The result shows that Downtown Toronto and evenings & noon have relatively high occurrences of crimes, which could be used as suggestions for both individual and government decision-making."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

“Crime is a complicated topic and there are few easy answers.” [@newark2011canadian] Major crimes statistics is an important tool for resource allocation and law enforcement strategy. It also provides the public a clear picture of the safety and helps individuals make informed decisions, influencing choices such as when to go out and where to buy a house. For this paper, I will use open-access data, titled ‘Major Crime Indicator’ from the OpenDataToronto [@rOpenDataToronto] to analyze the distribution of major crimes—classified as Assault, Break and Enter, Auto Theft, Robbery, and Theft Over—over the past ten years. My analysis will also delve into practical actions ordinary citizens and the government can do to enhance safety and efficiency.

@@

## Data Collection
The dataset contains all Major Crime Indicators occurrences by report date. The categories of major crimes are Assault, Break and Enter, Auto Theft, Robbery and Theft Over. The latest refresh was on Jan 11, 2024.

@@

## Data Analysis and Visualizations
The Major Crimes in Toronto dataset contains 27 columns and 372899 rows, including the repeated occurrences with different crime categories. All the samples of major crimes were reported from 2014 to 2023, and each sample includes ID, report time, occurrence time, police division, location type,  neighbourhood, and premise type. Among 27 columns, the variables I will use are x_id, report_year, occ_year, occ_month, occ_dow (day of the week offence occurred), occ_hour, mci_category and division. I will create two more variables, division_area, which put 18 police divisions (including a NSA) into 6 regions of Toronto (including Other for NSA), and weekend_weekday, which indicates whether the occurrence happen on a weekday or a weekend.

```{r}
#| message: false
#| echo: false
#| label: tbl-clean_mci_data
#| tbl-cap: The Number of Major Crimes occurred and the Month with the Most Crimes (2014-2023)
# Create a table with year, 2014 to 2023, the number of crimes in each year, and the month with the most crimes correspondingly.
clean_mci_data |>
  filter(occ_year >= 2014) |>
  group_by(occ_year, occ_month) |>
  summarise(n = n()) |>
  group_by(occ_year) |>
  summarise(
    total_number_of_crimes = sum(n),
    the_most_frequent_month = occ_month[which.max(n)]
  ) |>
  distinct(occ_year, .keep_all = TRUE) |>
  select(occ_year, total_number_of_crimes, the_most_frequent_month) |> 
  kable(
    col.names = c("Year", "Total Number of Crimes", "Month with the Most Crimes"),
    booktabs = TRUE
  )
```
Table 1 @ref(tbl-clean_mci_data) provides a brief summary of the total number of crimes from 2014 to 2023. Over this period, the overall number of crimes shows a consistent increase, except for the years 2020 and 2021, which have a decline probably due to the impact of the pandemic. However, in 2022 and 2023, there is a significant increase by more than 6000. In each year, the month with the highest number of crimes tends to be either July, August, or October, concentrating primarily in the summer-fall period.

@@

Figure 1 @ref(fig-types_of_crimes) visualizes the annual distribution of major crime types with five categories: Assault, Break and Enter, Auto Theft, Robbery, and Theft Over. Across all years, 'Assault' records the highest count, followed by 'Break and Enter' and 'Auto Theft'. 'Robbery' and 'Theft Over' show comparatively lower numbers. Notably, there has been a slight decrease in “Break and Enter” since the pandemic, and an obvious increase in 'Auto Theft' during 2022 and 2023, leading ‘Auto Theft’ surpassing the number of 'Break and Enter’.

@@

```{r}
#| include: false
clean_mci_data <- clean_mci_data |> 
  mutate(
    weekday_weekend =
      case_match(
        occ_dow,
        "Wednesday" ~ "Weekday",
        "Tuesday" ~ "Weekday",
        "Thursday" ~ "Weekday",
        "Sunday" ~ "Weekend",
        "Monday" ~ "Weekday",
        "Friday" ~ "Weekday",
        "Saturday" ~ "Weekend",
        "" ~ "Other"))
clean_mci_data
```
There are 18 police divisions, which are D11, D12, D13, D14, D22, D23, D31, D32, D33, D41, D42, D43, D51, D52, D53, D54, D55, and NSA. I put them into 6 geographical regions, Etobicoke-York, Etobicoke, North York, Scarborough, Toronto-East York, and other as shown in Table 2 @ref(tbl-division_area). Based on Figure 2 @ref(fig-area_of_crimes), Toronto-East York reports the highest amount of crimes, and Etobicoke-York has the lowest. North York, Scarborough, and Etobicoke-York have similar numbers, but the rate of increase in North York is higher than that observed in the other two. It is crucial to recognize that, though Toronto-East York records the most crimes, it also has the largest number of residents. Therefore, it could be biased to claim that Toronto-East York has the highest crime rate, or North York is getting dangerous, as North York has witnessed a growing population in recent years. Moreover, when creating this dataset, one occurrence number may have a few rows with different major crime indicators types used to categorize the occurrence. Therefore, some incidents of 'Auto Theft,' 'Robbery,' and 'Theft Over' may be potentially recorded multiple times under 'Break and Enter.' Consequently, the number of crimes classified as just ‘Break and Enter’ is overestimated, the actual number may be lower than what is shown now in the dataset.
```{r}
#| message: false
#| echo: false
#| label: tbl-division_area
#| tbl-cap: Police divisions classified into six areas (including other)

@@

```{r}
#| message: false
#| echo: false
#| label: fig-area_of_crimes
#| fig-cap: Total Number of Crimes Occurred in Different Police Division Areas (2014-2023)
clean_mci_data |>
  filter(occ_year >= 2014) |>
  count(occ_year, division_area) |>
  ggplot(aes(x = occ_year, y = n, color = division_area)) +
  geom_point() +
  geom_jitter() +
  geom_smooth(method = lm, se = FALSE) +
  labs(x = "Year",
       y = "Total Number of Crimes",
       color = "Division Area")+
  scale_x_continuous(breaks = seq(2014, 2023, by = 1))
```

```{r}
#| message: false
#| echo: false
#| label: fig-time_of_crimes
#| fig-cap: Total Number of Crimes Occurred in a Day
clean_mci_data |>
  filter(weekday_weekend == 'Weekday' | weekday_weekend == 'Weekend') |>
  count(occ_hour, weekday_weekend) |>
  ggplot(mapping = aes(x = occ_hour, y = n, color = weekday_weekend)) +
  geom_line() +
  theme(legend.position = "right") +
  labs(x = "Time In a Day", y = "Number of Crimes", color = "Weekday or Weekend") +
  scale_color_manual(values = c("Weekday" = "steelblue3", "Weekend" = "indianred3"))+
  scale_x_continuous(breaks = seq(0, 23, by = 1))
```

Figure 3 @ref(fig-time_of_crimes) indicates the preferred time to commit crimes. The graph reveals a higher likelihood of criminal behaviours during the evening, with occurrences steadily increasing as the night progresses towards midnight and afterwards, the occurrences decrease until six in the morning. Interestingly, there is a surprising local peak at noon, while 11:00 and 13:00 do not show a high probability.

The dataset named “Police Annual Statistical Report - Reported Crimes” from OpenDataToronto also provided crime data, including crime id, report_year, division, category, subtype and count. However, it overlaps with the information available in “Major Crime Indicators” dataset. Given that “Major Crime Indicators” contains more detailed records, I only used “Major Crime Indicators” in this paper.

@@

## Conclusion
In summary, ‘Assault’ has always ranked as the most common major crime over the past ten years. However, the difference between ‘Auto Theft’ and ‘Assault’ is shrinking, as the number of ‘Auto Theft’ is increasing rapidly. There need be more safety actions on automobiles, and people should lock the doors and not leave any packages visible through car windows to improve automobile security. Due to the difficulty to prove ‘Assault’ as a crime, installing cameras at the front door or home security systems is a good idea to provide evidence for ‘Assault’, while it can also record suspicious individuals and vehicles near one's residence. Taking proactive steps, such as responding to knocks cautiously while keeping the door locked, and refraining from touching anything until the police have inspected, is crucial for personal safety and should be instilled in children as well.

@@

Focus on safety during midnight and noon hours is important too. The most common time for a crime occurrence include noon and midnight, especially between 6:00 PM to 23:00 PM. In order to prevent such incidents, people who work from home but go out for lunch, as well as those with children returning home for lunch around noon, should keep their ears and eyes open for suspicious surroundings. Evening hours in any region also need more cautions from residents and security from the government.
